---
title: "birdsfirefrequency1"
date: "5/07/2021"
output:
  word_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---
  
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(here)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(runjags)
library(gridExtra)

# Set to TRUE to force re-fitting of models or FALSE to load
# previously saved outputs
REFIT_MODEL <- FALSE

# Cache directory for models
MODEL_DIR <- here("models")
if (!dir.exists(MODEL_DIR)) {
  dir.create(MODEL_DIR)
}

# convenience function to load contents of an RData file
# into a named object
load_from <- function(.path) {
  load(.path)
  objs <- ls(all.names = FALSE)
  get(objs[1], inherits = FALSE)
}

# ggplot default theme
theme_set( theme_bw() )

```  


## Avian movement groups and forest fire frequency

This document contains code written by Michael Bedward and Michael Franklin for the multispecies occupancy model presented in:  
Michael J. M. Franklin, Richard E. Major, Michael Bedward, and Ross A. Bradstock.  
'Relative avian mobility linked to use of fire-affected resources in forested landscapes'.  
Accepted for publication in Forest Ecology and Management on 23/06/2021.  

### Predictor variables
*Occupancy submodel:*  
MFC (fire frequency) was the area-weighted mean number of fires 1971-2002 over each 33 ha site.  
L1785 was the presence (1) or absence (0) of long-unburnt dry sclerophyll forest (> 5 ha) external to the site, but within a radius of 1,785 m of the site centre (10 km^2).   
NPL established whether sites were on the Newnes Plateau (NPL = 1) or Blue Mountains (NPL = 0).    
*Detection submodel:*    
ESD was the standard deviation of elevation over each site.        
  
### Bird data  
Survey data are presence or absence of species for each site and replicate survey session.  

```{r}

##### read in the bird data
path <- here("data", "all_records.csv")
DAT.survey <- read.csv(path, stringsAsFactors = FALSE)

# Species codes, names and info (note, 74 of the 86 listed species were recorded in these sites)
path <- here("data", "species_lookup2.csv")

SpeciesLookup <- read.csv(path, stringsAsFactors = FALSE) %>%
  # Reduce to species present in the survey data
  dplyr::filter(speciescode %in% DAT.survey$speciescode) %>%
  
  # Add an integer index for model use and discard existing index var
  select(-species) %>%
  arrange(speciescode) %>%
  mutate(species = row_number())

# For each species and site, sum the number of reps in which a detection was recorded
dat.counts <- DAT.survey %>%
  group_by(site, speciescode) %>%
  summarize(nobs = sum(observed)) %>%
  ungroup() %>%

  # Subset to species in SpeciesLookup
  filter(speciescode %in% SpeciesLookup$speciescode) %>%

  # Convert to wide format
  tidyr::spread(speciescode, nobs) %>%
  
  # Ensure site order
  arrange(site) %>%
  select(-site) %>%
  as.matrix()

```

Prior incidence groups for species. The suffix 'H' in 'incidenceH' and 'incidencecodeH' below denotes the refined (by habitat) version of the incidence groups for species priors.

```{r}

incidenceHLevels <- SpeciesLookup %>%
  group_by(incidenceH, incidencecodeH) %>%
  summarize(nspecies = n()) %>%
  arrange(incidenceH)

SpeciesLookup <- SpeciesLookup %>%
  mutate(incidencecodeH = factor(incidencecodeH, levels = incidenceHLevels$incidencecodeH))

```

Site attributes.

```{r}

path <- here("data", "sites_lookup2.csv")

SitesLookup <- read.csv(path, stringsAsFactors = FALSE) 

```


## Model G

### JAGS code

```{r}

model_G.code <- "model {
  for (site in 1:nsites) {
    for (sp in 1:nspp) {
      # Model the observations
      num.obs[site,sp] ~ dbin(present[site,sp] * p.detect[site,sp], nreps[site])

      # Model occupancy: the true (unknown) presences
      present[site,sp] ~ dbern(p.present[site,sp])

      logit(p.present[site, sp]) <- 
        beta.sp[sp] +
        beta.site[site] +
        beta.NPL * NPL[site] +
        beta.MFC[sp] * MFC[site] +
        beta.L1785[sp] * L1785pa[site]
    }

    beta.site[site] ~ dnorm(0, sd.site^(-2))
  }
  
  # Standard deviation of site-level random effects
  sd.site ~ dexp(1)
  
  # Area effect for NPL
  beta.NPL ~ dnorm(0, 1)

##################

  for (sp in 1:nspp) {
    # The intercept for each species (beta.sp) is drawn from a Normal
    # distribution for the corresponding prior incidence group with
    # mean mu.sp.grp[grp] and common standard deviation sd.sp
    #
    beta.sp[sp] ~ dnorm(mu.sp.grp[incidgrp[sp]], sd.sp^(-2))
  }

  # Standard deviaion of the species intercepts.
  sd.sp ~ dexp(1)
  
  # Priors for incidenceH group means ('mu.sp.grp')
  for (i in 1:nincidgrps) {
    unordered.sp.grp[i] ~ dnorm(beta0, 0.1)
  }
  mu.sp.grp <- sort(unordered.sp.grp)

  # Grand intercept
  beta0 ~ dnorm(0, 1)
  
######################

  # Weighted mean fire count effect. 
  # The parameter value for each species is drawn from a distribution
  # for the corresponding mobility group.
  for (sp in 1:nspp) {
    beta.MFC[sp] ~ dnorm(mu.MFC.mobility[mobility[sp]], 
                         sd.MFC^(-2))
  }
  sd.MFC ~ dexp(1)
  
  # Group-level means
  for (grp in 1:nmobility) {
    mu.MFC.mobility[grp] ~ dnorm(0, 1)
  }
  
######################

  # Effect of presence of long-unburnt dry sclerophyll forest within site radius.
  # The parameter value for each species is drawn from a distribution
  # for the corresponding mobility group.
  # 
  for (sp in 1:nspp) {
    beta.L1785[sp] ~ dnorm(mu.L1785.mobility[mobility[sp]], sd.L1785^(-2))
  }
  sd.L1785 ~ dexp(1)
  
  # Group-level means
  for (grp in 1:nmobility) {
    mu.L1785.mobility[grp] ~ dnorm(0, 1)
  }
  
######################

# Detection model

  for (sp in 1:nspp) {
    for (site in 1:nsites) {
      # A general level of detectability for the species plus the modifying,
      # species-specifc influence of elevation variability
      
      logit(p.detect[site, sp]) <- gamma.sp[sp] + gamma.ESD[sp]*ESD[site]
    }
    
    gamma.sp[sp] ~ dnorm(gamma.sp0, 1)
    gamma.ESD[sp] ~ dnorm(gamma.ESD.0, 1)
  }
  
  gamma.sp0 ~ dnorm(0, 1)
  gamma.ESD.0 ~ dnorm(0, 1)
}"

```


### Model data

```{r}

nsites <- nrow(dat.counts)
nspp <- ncol(dat.counts)

dat.preds <- SitesLookup %>%
  # Ensure we have site order
  arrange(site) %>%
  
  select(site, NPL, MFC, L1785, ESD) %>%
  
  mutate(ESDs = scale(ESD)[,1],
         MFCs = scale(MFC)[,1])

# Mobility group as a factor
fmobility = factor(SpeciesLookup$mobility2, 
                   levels = c("Migratory", "Nomadic", "Sedentary"))


model_G.data <- list(
  nsites = nsites,
  nspp = nspp,
  nreps = rep(10, nsites),
  num.obs = dat.counts,
  
  nincidgrps = nrow(incidenceHLevels),
  incidgrp = SpeciesLookup$incidenceH,
  
  NPL = dat.preds$NPL,
  ESD = dat.preds$ESDs,
  MFC = dat.preds$MFCs,
  L1785pa = ifelse(dat.preds$L1785 > 0, 1, 0),
  
  # mobility group for each species
  mobility = fmobility,
  
  nmobility = nlevels(fmobility)
)

```


### Run the model

```{r}

model.path <- file.path(MODEL_DIR, "model_G_allspp.RData")

if (REFIT_MODEL || !file.exists(model.path)) {
  model_G <- run.jags(
    model = model_G.code, 
    monitor = c("beta0", 
                "beta.sp", "mu.sp.grp", "sd.sp", 
                "gamma.sp",
                "beta.site", "sd.site",
                "beta.NPL",
                "beta.MFC",
                "beta.L1785",
                "mu.MFC.mobility",
                "mu.L1785.mobility"), 
    
    n.chains = 4, 
    data = model_G.data, 
    
    inits = function() {
      list(beta0 = 0,
           beta.sp = runif(nspp),
           beta.site = rep(0, nsites),
           beta.NPL = 0,
           mu.MFC.mobility = rep(0, nlevels(fmobility)),
           mu.L1785.mobility = rep(0, nlevels(fmobility)),
           gamma.sp = runif(nspp),
           gamma.ESD = rep(0, nspp))
    }, 
    
    method = "parallel", 
    sample = 1000,         
    thin = 25,            
    burnin = 1000)
  
  save(model_G, file = model.path)

} else {
  # load previously cached model object
  model_G <- load_from(model.path)
}

```


### Check model convergence

```{r}

x <- summary(model_G)

cat("Min effective samples:", min(x[, "SSeff"]), "\n")
max.gelman <- cat("Max Gelman-Rubin value:", max(x[, "psrf"]), "\n")

```
